---
title: "Analysis pre-analytical plasma and serum - MOFA"
author: "Thomas Naake"
date: "05/20/2021"
fig_width: 15
fig_height: 10
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: TRUE
    theme: united
    number_sections: true
    highlight: tango
    self_contained: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf")
knitr::opts_knit$set(root.dir = "C:/Users/naake/Documents/GitLab/20210520_Proteomics_Metabolomics_plasma_serum_MOFA/")
```


# Load the SummarizedExperiment objects

Load the `SummarizedExperiment` objects from metabolomics and proteomics.

```{r read_RDS_files}
## load protein SE
path_to_file <- "C:/Users/naake/Documents/GitLab/20210517_Proteomics_Torsten_plasma_serum_proteins/Preanalytical_SummarizedExperiment_Protein_transformed.RDS"
protein <- readRDS(path_to_file)

## load metabolite SE
path_to_file <- "C:/Users/naake/Documents/GitLab/20210112_Metabolomics_Hagen_plasma_serum_metabolites/Preanalytical_SummarizedExperiment_Metabolite_transformed.RDS"
metabolite <- readRDS(path_to_file)
```

Create same columns. First, for `protein` calculate means of replicates.
```{r remove_and_rename, echo = TRUE, eval = TRUE, message = FALSE}
library(SummarizedExperiment)
prot_name <- colData(protein)$name
prot_name <- unlist(
    lapply(stringr::str_split(prot_name, pattern = "_Slot1"), "[", 1))
prot_name_dup <- prot_name[duplicated(prot_name)]

## rename the columns in protein
colnames(protein) <- prot_name
colData(protein)$name <- prot_name

## iterate trough the duplicated columns and calculate the mean between 
## replicates
for(i in seq_along(prot_name_dup)) {
    a_dup <- assay(protein)[, prot_name_dup[i] == colnames(protein)]
    a_dup_mean <- apply(a_dup, 1, mean)
    assay(protein)[, prot_name_dup[i] == colnames(protein)] <- a_dup_mean
}

## remove the duplicated columns from protein
protein <- protein[, !duplicated(prot_name)]

## remove the Biobank samples 
protein <- protein[, !grepl(colnames(protein), pattern = "Biobank")]
```

Second, remove the columns with `C2` centrifugation and 
rename the colnames in `metabolites` (Type_Temp_Time_Person), 


```{r renaming}
## remove the C2 samples
metabolite <- metabolite[, !colData(metabolite)$centr == "C2"]

## rename the colnames
colnames(metabolite) <- paste0(
    colData(metabolite)$tissue, "_", colData(metabolite)$temp, "_",
    colData(metabolite)$time, "h_", colData(metabolite)$person)
colData(metabolite)$name <- colnames(metabolite)
```

# Create MOFA object

```{r load_data, echo = TRUE, eval = TRUE, message = FALSE}
library(MOFA2)
a_protein <- assay(protein)
a_metabolite <- assay(metabolite)

## what are the missing samples 
## samples in protein, but not in metabolite
colnames(a_protein)[!colnames(a_protein) %in% colnames(a_metabolite)]
## samples in metabolite, but not in protein
colnames(a_metabolite)[!colnames(a_metabolite) %in% colnames(a_protein)]

## create intersect and truncate the assay that they contain the sampe samples
cols <- intersect(colnames(a_protein), colnames(a_metabolite))

data <- list(metabolite = assay(metabolite)[, cols], 
             protein = assay(protein)[, cols])

## remove the feature that only contains NA values
data[["protein"]] <- data[["protein"]][!apply(data[["protein"]], 1, 
                                                function(x) all(is.na(x))), ]

## adjust the sample metadata
samp <- cols
metadata_df <- data.frame(
    sample = samp,
    group = unlist(lapply(strsplit(samp, split = "_"), "[", 1)),
    temperature = unlist(lapply(strsplit(samp, split = "_"), "[", 2)),
    time = unlist(lapply(strsplit(samp, split = "_"), "[", 3)),
    person = unlist(lapply(strsplit(samp, split = "_"), "[", 4))
)

## cretae the MOFA object from the matrices
MOFA <- create_mofa(data, groups = metadata_df$group)

data_opts <- get_default_data_options(MOFA)
##data_opts$scale_views <- TRUE
##data_opts$scale_groups <- TRUE
model_opts <- get_default_model_options(MOFA)
model_opts$num_factors <- 15
train_opts <- get_default_training_options(MOFA)
train_opts$maxiter <- 10000
## drop_factor_threshold: numeric indicating the threshold on fraction of 
## variance explained to consider a factor inactive and drop it from the 
## model. a value of 0.01 implies that factors explaining less 
## than 1% of variance (in each view) will be dropped
train_opts$seed <- 42
train_opts$drop_factor_threshold <- 0.01
train_opts$convergence_mode <- "slow"

MOFA <- prepare_mofa(
    object = MOFA,
    data_options = data_opts,
    model_options = model_opts,
    training_options = train_opts
)

samples_metadata(MOFA) <- metadata_df

plot_data_overview(MOFA)
```

```{r run_mofa, eval = TRUE, echo = TRUE}
## finally train the model
MOFA_trained <- run_mofa(MOFA)
saveRDS(MOFA_trained, file = "MOFA_trained_proteomics_metabolomics.RDS")
```

```{r, eval = TRUE, echo = FALSE}
MOFA_trained <- readRDS("MOFA_trained_proteomics_metabolomics.RDS")
```

Check if there is correlation between the factors.
Factors should not show any significant correlation.
```{r plot_factor_cor}
plot_factor_cor(MOFA_trained)
```



# Factor association analysis

Let's compute the interconnection of the sample metadata and factor values 
before zooming into the individual factors.

```{r eval=TRUE, echo=FALSE}
correlate_factors_with_covariates <- function (object, covariates, factors = "all", groups = "all", 
    abs = FALSE, plot = c("log_pval", "r"), alpha = 0.05, 
    return_data = FALSE, transpose = FALSE, ...) {
    if (!is(object, "MOFA")) 
        stop("'object' has to be an instance of MOFA")
    groups <- MOFA2:::.check_and_get_groups(object, groups)
    metadata <- samples_metadata(object)
    metadata <- metadata[metadata$group %in% groups, ]
    if (is.character(covariates)) {
        stopifnot(all(covariates %in% colnames(metadata)))
        covariates <- metadata[, covariates, drop = FALSE]
    }
    else if (is.data.frame(covariates)) {
        samples <- metadata$sample
        if (is.null(rownames(covariates))) 
            stop("The 'covariates' data.frame does not have samples names")
        stopifnot(all(rownames(covariates) %in% samples))
        covariates <- metadata[match(rownames(covariates), metadata$sample), 
            ]
    }
    else {
        stop("covariates argument not recognised. Please read the documentation: ?correlate_factors_with_covariates")
    }
    cols <- which(sapply(covariates, is.character))
    if (length(cols >= 1)) {
        covariates[cols] <- lapply(covariates[cols], as.factor)
    }
    cols <- which(!sapply(covariates, class) %in% c("numeric", 
        "integer"))
    if (length(cols >= 1)) {
        cols.factor <- which(sapply(covariates, class) == "factor")
        covariates[cols] <- lapply(covariates[cols], as.numeric)
        warning("There are non-numeric values in the covariates data.frame, converting to numeric...")
        covariates[cols] <- lapply(covariates[cols], as.numeric)
    }
    stopifnot(all(sapply(covariates, class) %in% c("numeric", 
        "integer")))
    factors <- MOFA2:::.check_and_get_factors(object, factors)
    Z <- get_factors(object, factors = factors, groups = groups, 
        as.data.frame = FALSE)
    Z <- do.call(rbind, Z)
    cor <- psych::corr.test(Z, covariates, method = "pearson", 
        adjust = "BH")
    plot <- match.arg(plot)
    if (plot == "r") {
        stat <- cor$r
        if (abs) 
            stat <- abs(stat)
        if (transpose) 
            stat <- t(stat)
        if (return_data) 
            return(stat)
        corrplot(stat, tl.col = "black", title = "Pearson correlation coefficient", 
            ...)
    }
    else if (plot == "log_pval") {
        stat <- cor$p
        stat[stat > alpha] <- 1
        if (all(stat == 1)) 
            stop("All p-values are 1.0, nothing to plot")
        stat <- -log10(stat)
        stat[is.infinite(stat)] <- 1000
        if (transpose) 
            stat <- t(stat)
        if (return_data) 
            return(stat)
        col <- colorRampPalette(c("#FFFFFF", "#8B0000"))(100)
        pheatmap::pheatmap(stat, main = "log10 adjusted p-values", 
            cluster_rows = FALSE, color = col, ...)
    }
    else {
        stop("'plot' argument not recognised. Please read the documentation: ?correlate_factors_with_covariates")
    }
}
```

```{r correlate_factors_with_covariates}
correlate_factors_with_covariates(MOFA_trained,
                    covariates = c("group", "person", "temperature", "time"),
                    plot = "log_pval")
```

Factors 1, 3, and 4 seem to be associated with `person` and
factor 8 with `time`

# Variance decomposition

The first step in the MOFA analysis is to quantify the amount of variance 
explained (R2) by each factor in each data modality.
```{r total_variance}
## total variance explained per view and group
MOFA_trained@cache[["variance_explained"]]

## variance explained for every factor in per view and group
head(MOFA_trained@cache$variance_explained$r2_per_factor[[1]]) # group 1 == plasma
head(MOFA_trained@cache$variance_explained$r2_per_factor[[2]]) # group 2 == serum

## plot variance explained estimates
plot_variance_explained(MOFA_trained, x = "view", y = "factor") + 
    ggplot2::scale_fill_gradient(low = "#FFFFFF", high = "#8B0000")
```

Plot that shows the explained variation per data modality
```{r variance_explained}
plot_variance_explained(MOFA_trained, x = "group", y = "factor", 
    plot_total = T)[[2]]
```

## Visualization of single factors in latent space

Each MOFA factor captures a different dimension of heterogeneity in the data.
Mathematically, each factor ordinates cells along a one-dimensional axis 
centered at zero. A Factor is a linear combination of initial features and 
represents a source of data variability.
Samples with different signs manifest opposite phenotypes 
along the inferred axis of variation, with higher absolute value indicating 
a stronger effect. On the image below, Factor values are located around the 
zero axis as the data is mandatory centered prior to the Factor analysis.

Note that the interpretation of factors is analogous to 
the interpretation of the principal components in PCA. A larger absolute value 
correlates with a more expressed biological effect.

```{r vis_singlefactor_latentspace}
## for factor 1 (separation along person)
p1 <- plot_factor(MOFA_trained, factor = 1, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 1, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 2 (separation along person)
p1 <- plot_factor(MOFA_trained, factor = 2, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 2, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 3 (separation along person)
p1 <- plot_factor(MOFA_trained, factor = 3, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 3, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 4 (separation along person)
p1 <- plot_factor(MOFA_trained, factor = 4, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 4, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 5
p1 <- plot_factor(MOFA_trained, factor = 5, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 5, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 6
p1 <- plot_factor(MOFA_trained, factor = 6, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 6, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 7
p1 <- plot_factor(MOFA_trained, factor = 7, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 7, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)

## for factor 8 (separation along time)
p1 <- plot_factor(MOFA_trained, factor = 8, color_by = "person", 
                  shape_by = "time")
p2 <- plot_factor(MOFA_trained, factor = 8, color_by = "person", 
                  shape_by = "temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Visualization of combinations of factors
Scatter plots

```{r vis_comb_factors}
plot_factors(MOFA_trained, factors = 1:4, color_by = "person")
```

## Visualization of feature weights

The weights provide a score for how strong each feature relates to each factor.
Features with no association with the factor have values close to zero, while 
features with strong association with the factor have large absolute values. 
The sign of the weight indicates the direction of the effect: a positive weight
indicates that the feature has higher levels in the cells with positive factor 
values, and vice versa.

Weights can be plotted using `plot_weights` (beeswarm plots) or
`plot_top_weights` (scatter plots).

```{r plot_weights}
## axis that explains most of the variance, factor 1
plot_weights(MOFA_trained, view = "metabolite", factor = 1,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "metabolite", factor = 1, nfeatures = 10)

plot_weights(MOFA_trained, view = "protein", factor = 1,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "protein", factor = 1, nfeatures = 10)

## factor 2
plot_weights(MOFA_trained, view = "metabolite", factor = 2,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "metabolite", factor = 2, nfeatures = 10)

plot_weights(MOFA_trained, view = "protein", factor = 2,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "protein", factor = 2, nfeatures = 10)

## factor 3
plot_weights(MOFA_trained, view = "metabolite", factor = 3,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "metabolite", factor = 3, nfeatures = 10)

plot_weights(MOFA_trained, view = "protein", factor = 3,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "protein", factor = 3, nfeatures = 10)

## factor 4
plot_weights(MOFA_trained, view = "metabolite", factor = 4,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "metabolite", factor = 4, nfeatures = 10)

plot_weights(MOFA_trained, view = "protein", factor = 4,
    nfeatures = 10, ## Number of features to highlight
    scale = TRUE, ## Scale weights from -1 to 1
    abs = FALSE) ## Take the absolute value?
plot_top_weights(MOFA_trained, view = "protein", factor = 4, nfeatures = 10)

## axis that associates with time
## factor 8
plot_weights(MOFA_trained, view = "metabolite", factor = 8, nfeatures = 10, 
    scale = TRUE, abs = FALSE)
plot_top_weights(MOFA_trained, view = "metabolite", factor = 8, nfeatures = 10)

plot_weights(MOFA_trained, view = "protein", factor = 8, nfeatures = 10,
    scale = TRUE, abs = FALSE)
plot_top_weights(MOFA_trained, view = "protein", factor = 8, nfeatures = 10)
```

# Visualization of patterns in the input data

Instead of looking at weights, it is useful to observe the coordinated 
heterogeneity that MOFA captures in the original data. This can be done 
using the `plot_data_heatmap` and `plot_data_scatter` function.

## Heatmaps

Heatmap of observations. Top features are selected by its weight in the 
selected factor. By default, samples are ordered according to their 
corresponding factor value.

```{r plot_data_heatmap}
## for the person-associated factors
## factor 1
plot_data_heatmap(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 1, ## factor of interest
  features = 10, ## number of features to plot (they are selected by weight)
  ## extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = TRUE, 
  show_colnames = TRUE, scale = "row", 
  annotation_samples = "person"
)

## factor 2
plot_data_heatmap(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 2, ## factor of interest
  features = 10, ## number of features to plot (they are selected by weight)
  ## extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = TRUE, 
  show_colnames = TRUE, scale = "row",
  annotation_samples = "person"
)

## factor 3
plot_data_heatmap(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 3, ## factor of interest
  features = 10, ## number of features to plot (they are selected by weight)
  ## extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = TRUE, 
  show_colnames = TRUE, scale = "row",
  annotation_samples = "person"
)

## factor 4
plot_data_heatmap(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 4, ## factor of interest
  features = 10, ## number of features to plot (they are selected by weight)
  ## extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = TRUE, 
  show_colnames = TRUE, scale = "row",
  annotation_samples = "person"
)

## for time-associated factor 8
plot_data_heatmap(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 8, ## factor of interest
  features = 10, ## number of features to plot (they are selected by weight)
  ## extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = TRUE, 
  show_colnames = TRUE, scale = "row",
  annotation_samples = "time"
)
```

## Scatter plots

Scatter plots of observations vs factor values. It is useful to add a linear
regression estimate to visualize if the relationship between (top) features and
factor values is linear.

```{r plot_data_scatter}
plot_data_scatter(MOFA_trained,
  view = "metabolite", ## view of interest
  factor = 8, ## factor of interest
  features = 5, ## number of features to plot (they are selected by weight)
  add_lm = TRUE, ## add linear regression
  color_by = "time"
)
```

## Non-linear

The MOFA factors are linear (as in Principal Component analysis). Nevertheless, 
the MOFA factors can be used as input to other methods that learn compact 
nonlinear manifolds (t-SNE or UMAP).

Run UMAP and t-SNE.

```{r tsne}
set.seed(42)
MOFA_trained <- run_umap(MOFA_trained)
MOFA_trained <- run_tsne(MOFA_trained, perplexity = 4)

plot_dimred(MOFA_trained,
  method = "TSNE", ## method can be either "TSNE" or "UMAP"
  color_by = "person"
)
```
