---
title: "Analysis pre-analytical plasma and serum - Limma to test person effect"
author: "Thomas Naake"
date: "05/20/2021"
fig_width: 15
fig_height: 10
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: TRUE
    theme: united
    number_sections: true
    highlight: tango
    self_contained: no
---

A more head-on, direct analysis to analyse individual effects simply looks at
differentially quantified/abundant features with a tool such as limma (either covariate by covariate, or also jointly) and counting those with, say, 
FDR <0.1 as a measure of how important the covariate is.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf")
knitr::opts_knit$set(root.dir = "C:/Users/naake/Documents/GitLab/20210520_Proteomics_Metabolomics_plasma_serum_MOFA/")
```

# Load packages

```{r}
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(limma))
```

# Load the SummarizedExperiment objects

Load the `SummarizedExperiment` objects from metabolomics and proteomics.

```{r read_RDS_files}
## load protein SE
path_to_file <- "C:/Users/naake/Documents/GitLab/20210517_Proteomics_Torsten_plasma_serum_proteins/Preanalytical_SummarizedExperiment_Protein_imputed.RDS"
protein <- readRDS(path_to_file)
protein <- protein[, protein$Replicate != "Biobank"]
protein_s <- protein[, protein$Type == "Serum"]
protein_p <- protein[, protein$Type == "Plasma"]

## load metabolite SE
path_to_file <- "C:/Users/naake/Documents/GitLab/20210112_Metabolomics_Hagen_plasma_serum_metabolites/Preanalytical_SummarizedExperiment_Metabolite_imputed.RDS"
metabolite <- readRDS(path_to_file)
metabolite$person <- factor(metabolite$person,
    levels = as.character(unique(metabolite$person)))
metabolite <- metabolite[, metabolite$centr == "C1"]
metabolite_s <- metabolite[, metabolite$tissue == "Serum"]
metabolite_p <- metabolite[, metabolite$tissue == "Plasma"]

```

```{r}
p <- 0.05 
n <- Inf

## protein
mM_p <- model.matrix(as.formula("~ 1 + Replicate + Temperature * Time"), 
    data = colData(protein_p))
fit <- lmFit(assay(protein_p), design = mM_p)
fit2 <- eBayes(fit, robust = TRUE)
tT <- topTable(fit2, coef = grep("Replicate", colnames(coef(fit))), 
    p.value = p, number = n)
nrow(tT)
rmarkdown::paged_table(tT)

mM_s <- model.matrix(as.formula("~ 1 + Replicate + Time"), 
    data = colData(protein_s))
fit <- lmFit(assay(protein_s), design = mM_s)
fit2 <- eBayes(fit, robust = TRUE)
tT <- topTable(fit2, coef = grep("Replicate", colnames(coef(fit))), 
    p.value = p, number = n)
nrow(tT)
rmarkdown::paged_table(tT)


## metabolite
mM_p <- model.matrix(as.formula("~ 1 + person + temp * time"), 
    data = colData(metabolite_p))
fit <- lmFit(assay(metabolite_p), design = mM_p)
fit2 <- eBayes(fit, robust = TRUE)
tT <- topTable(fit2, coef = grep("person", colnames(coef(fit))), 
    p.value = p, number = n)
nrow(tT)
rmarkdown::paged_table(tT)

mM_s <- model.matrix(as.formula("~ 1 + person + time"), 
    data = colData(metabolite_s))
fit <- lmFit(assay(metabolite_s), design = mM_s)
fit2 <- eBayes(fit, robust = TRUE)
tT <- topTable(fit2, coef = grep("person", colnames(coef(fit))), 
    p.value = p, number = n)
nrow(tT)
rmarkdown::paged_table(tT)
```

