---
title: "Analysis pre-analytical plasma and serum"
author: "Thomas Naake"
date: "09/10/2021"
fig_width: 15
fig_height: 10
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: TRUE
    theme: united
    number_sections: true
    highlight: tango
    self_contained: no
---

```{r wd_metabolite, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf")
knitr::opts_knit$set(root.dir = "C:/Users/naake/Documents/GitLab/20210112_Metabolomics_Hagen_plasma_serum_metabolites/")
```

# Metabolites

## Load the data

Load packages.
```{r load_packages, message = FALSE, warning = FALSE}
library(MatrixQCvis)
library(ggplot2)
library(plotly)
```

Load the data. 
```{r load_metabolite, show = FALSE, eval = TRUE, echo = TRUE}
se_t <- readRDS("Preanalytical_SummarizedExperiment_Metabolite_transformed.RDS")
se_i <- readRDS("Preanalytical_SummarizedExperiment_Metabolite_imputed.RDS")

## create factors for the time to test between 0 h vs. 8 h 
colData(se_t)$time <- as.factor(colData(se_t)$time)
colData(se_i)$time <- as.factor(colData(se_i)$time)

## the SummarizedExperiment objects do not contain person1, refactor the person
## column
colData(se_t)$person <- factor(colData(se_t)$person, levels = 2:6)
colData(se_i)$person <- factor(colData(se_i)$person, levels = 2:6)
```

## Prepare data input for signature analysis (by Aurelien)

Perform for the plasma and serum samples separately a differential expression
analysis. Also, split prior to performing the analysis the data up in the 
different temperature regimes.

We will use here moderated t-tests adjusting for person-specific effects 
and check for the contrast `time` (time point `0h` vs `8h`).

### For plasma samples

Define the model matrix and contrasts.
```{r mM_plasma_metabolite}
se_p <- se_i[, colData(se_i)$tissue == "Plasma"]

## split up for temperature
se_p_4C <- se_p[, colData(se_p)$temp == "4C"]
se_p_RT <- se_p[, colData(se_p)$temp == "RT"]

## obtain the colData
cD_p_4C <- colData(se_p_4C)
cD_p_RT <- colData(se_p_RT)

## create the model matrix and define the contrast
mM_p_4C <- model.matrix(as.formula("~ 0 + person + time"), data = cD_p_4C)
contrasts_p_4C <- limma::makeContrasts(contrasts = "time8", levels = mM_p_4C)
mM_p_RT <- model.matrix(as.formula("~ 0 + person + time"), data = cD_p_RT)
contrasts_p_RT <- limma::makeContrasts(contrasts = "time8", levels = mM_p_RT)
```

Do the analysis for 4C.
```{r diff_ttest_plasma_metabolite, eval = TRUE}
## for 4C
fit <- limma::lmFit(assay(se_p_4C), design = mM_p_4C)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_p_4C)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_metabolites_time0vs8_plasma_4C.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```

Do the analysis for room temperature.
```{r diff_ttest_plasma_RT_metabolite, eval = TRUE}
## for RT
fit <- limma::lmFit(assay(se_p_RT), design = mM_p_RT)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_p_RT)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_metabolites_time0vs8_plasma_RT.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```

### For serum samples

For serum samples, there is only one temperature level, `RT`.

Define the model matrix and contrasts.
```{r mM_serum_metabolite}
se_s <- se_i[, colData(se_i)$tissue == "Serum"]

## split up for temperature
se_s_RT <- se_s[, colData(se_s)$temp == "RT"]

## obtain the colData
cD_s_RT <- colData(se_s_RT)

## create the model matrix and define the contrast
mM_s_RT <- model.matrix(as.formula("~ 0 + person + time"), data = cD_s_RT)
contrasts_s_RT <- limma::makeContrasts(contrasts = "time8", levels = mM_s_RT)
```

Do the analysis for room temperature.
```{r diff_ttest_serum_metabolite, eval = TRUE}
## for RT
fit <- limma::lmFit(assay(se_s_RT), design = mM_s_RT)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_s_RT)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_metabolites_time0vs8_serum_RT.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```


# Proteins

## Load the data


```{r setwd_protein, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/naake/Documents/GitLab/20210517_Proteomics_Torsten_plasma_serum_proteins")
```

Load the data. 
```{r load_protein, show = FALSE, eval = TRUE, echo = TRUE}
se_t <- readRDS("Preanalytical_SummarizedExperiment_Protein_transformed.RDS")
se_i <- readRDS("Preanalytical_SummarizedExperiment_Protein_imputed.RDS")

## remove the Biobank samples
se_t <- se_t[, se_t$Replicate != "Biobank"]
se_i <- se_i[, se_i$Replicate != "Biobank"]

## create factors for the time to test between 0 h vs. 8 h 
colData(se_t)$Time <- as.factor(colData(se_t)$Time)
colData(se_i)$Time <- as.factor(colData(se_i)$Time)

## refactor the person column
colData(se_t)$Replicate <- factor(colData(se_t)$Replicate, levels = 1:6)
colData(se_i)$Replicate <- factor(colData(se_i)$Replicate, levels = 1:6)
```

## Prepare data input for signature analysis (by Aurelien)

Perform for the plasma and serum samples separately a differential expression
analysis. Also, split prior to performing the analysis the data up in the 
different temperature regimes.

We will use here moderated t-tests adjusting for person-specific effects 
and check for the contrast `time` (time point `0h` vs `8h`).

### For plasma samples

Define the model matrix and contrasts.
```{r mM_plasma_protein}
se_p <- se_i[, colData(se_i)$Type == "Plasma"]

## split up for temperature
se_p_4C <- se_p[, colData(se_p)$Temperature == "4C"]
se_p_RT <- se_p[, colData(se_p)$Temperature == "RT"]

## obtain the colData
cD_p_4C <- colData(se_p_4C)
cD_p_RT <- colData(se_p_RT)

## create the model matrix and define the contrast
mM_p_4C <- model.matrix(as.formula("~ 0 + Replicate + Time"), data = cD_p_4C)
contrasts_p_4C <- limma::makeContrasts(contrasts = "Time8", levels = mM_p_4C)
mM_p_RT <- model.matrix(as.formula("~ 0 + Replicate + Time"), data = cD_p_RT)
contrasts_p_RT <- limma::makeContrasts(contrasts = "Time8", levels = mM_p_RT)
```

Do the analysis for 4C.
```{r diff_ttest_plasma_protein, eval = TRUE}
## for 4C
fit <- limma::lmFit(assay(se_p_4C), design = mM_p_4C)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_p_4C)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_proteins_time0vs8_plasma_4C.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```

Do the analysis for room temperature.
```{r diff_ttest_plasma_RT_protein, eval = TRUE}
## for RT
fit <- limma::lmFit(assay(se_p_RT), design = mM_p_RT)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_p_RT)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_proteins_time0vs8_plasma_RT.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```

### For serum samples

For serum samples, there is only one temperature level, `RT`.

Define the model matrix and contrasts.
```{r mM_serum_protein}
se_s <- se_i[, colData(se_i)$Type == "Serum"]

## split up for temperature
se_s_RT <- se_s[, colData(se_s)$Temperature == "RT"]

## obtain the colData
cD_s_RT <- colData(se_s_RT)

## create the model matrix and define the contrast
mM_s_RT <- model.matrix(as.formula("~ 0 + Replicate + Time"), data = cD_s_RT)
contrasts_s_RT <- limma::makeContrasts(contrasts = "Time8", levels = mM_s_RT)
```

Do the analysis for room temperature.
```{r diff_ttest_serum_protein, eval = TRUE}
## for RT
fit <- limma::lmFit(assay(se_s_RT), design = mM_s_RT)
fit <- limma::contrasts.fit(fit = fit, contrasts = contrasts_s_RT)
fit <- limma::eBayes(fit = fit, robust = TRUE)
tT <- limma::topTable(fit, number = Inf, adjust.method = "fdr", p.value = 1)
rmarkdown::paged_table(tT)
tT <- cbind(name = rownames(tT), tT)
partial_bundle(MatrixQCvis::volcanoPlot(tT))
write.table(tT, file = "diff_expr_ttest_proteins_time0vs8_serum_RT.txt", 
      sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```


